---
title: "Multivariate Cox regression analysis"
output: html_document
---

This template used to run multivariate cox regression analysis based on gene set score. This allows for controlling for confounders like tumor type, age, etc.

## Overall steps:

  1. Determine which clinical variables to control for- run univariate cox to identify variables that may have an effect on survival
  2. Determine baseline level of effect of gene set score on survival- univariate cox. Can compare to kaplan-meier survival results (should be very similar)
  3. Determine effect of gene set score on survival when clinical variables are included in model- multivariate cox
  
```{r}
library(survival)
library(survminer)
library(dplyr)

## Set working/script directory
base_dir = "/Users/jlabelle/"
working_dir<- paste0(base_dir, 
                     "Dropbox (Partners HealthCare)/FilbinLab/",
                     "data_analysis/Projects/git_repos/ProjectA/")
script_dir = "Dropbox (Partners HealthCare)/FilbinLab/data_analysis/Shared/Scripts/"

## Load helper functions
helper_functions<- paste0(base_dir, script_dir,"Survival_HelperFunctions.R")
source("~/Dropbox (Partners HealthCare)/Filbin lab/Shared/Scripts/Survival_HelperFunctions.R")

## Set analysis name
analysis_name<- "test"

## Set directories
data_dir<- paste0("analysis/KM_survival/", analysis_name, "/")
analysis_dir<- paste0("analysis/coxRegression/", analysis_name, "/")
figure_dir<- paste0("figures/coxRegression/", analysis_name, "/")
if(!dir.exists(figure_dir)){dir.create(figure_dir, recursive = TRUE)};if(!dir.exists(analysis_dir)){dir.create(analysis_dir)}

```

## Load in data
Input:
list of "FinalMeta" from SurvivalAnalysis_template.Rmd

Minimum columns required: 
  a. SampleID ("submitter_id") 
  b. time to death ("time") 
  c. alive/dead ("censored"; alive= F, dead=T). Should be set as logical. 
  d. At least 1 method for splitting samples. Can be 1) raw scores 2) high/low scores 3) raw proportion 4) high/low proportion. Each geneset/program should have its own column

2.Clinical metadata. Contains variables you want to control for- age/sex/diagnosis/treatment/etc. First column should be sample ID. 
Note: should remove any variables with NA values for any sample, or remove those samples.
```{r}
## Read in list with info for all cohorts tested
surv_obj_list<- readRDS(paste0(data_dir, "SurvivalInput_allCohorts.Rds"))

## Flatten FinalMeta into single df for all cohorts
final_meta_list<- lapply(surv_obj_list, function(x){
  tmp<- x$FinalMeta
  tmp$Cohort<- x$cohort
  return(tmp)
  })
FinalMeta<- do.call("rbind", final_meta_list)

## Set genelists/programs to test
genesets_test<- c("gluta", "gaba")
```


## Set variables to potentially control for
## These are variables from your clinical metadata that you think may have an effect on survival.
## Can also just use all variables
```{r}
## List with variables to control for (if running across multiple cohorts, may want to include "Cohort")
ToControlFor<- c("age_at_index", "gender", "race", "MyDiagnosis")

```


## Optional: 
1. Create additional variable, merging similar diagnoses together. This may not apply to all data.
2. Subset by diagnosis- bc glioblastoma diagnosis has such a large effect, may want to remove it from consideration
```{r}
FinalMeta <- meta
## Optional: add column to meta- astro/oligo/mixed/glio
FinalMeta$MyDiagnosis<-gsub("Mixed glioma", "MixedGlioma", gsub("Astrocytoma, NOS", "Astrocytoma", 
                         gsub("Oligodendroglioma, anaplastic", "Oligodendroglioma", 
                              gsub("Oligodendroglioma, NOS", "Oligodendroglioma", 
                                   gsub("Astrocytoma, anaplastic", "Astrocytoma", 
                                        gsub("Glioblastoma", "Glioblastoma",
                                             FinalMeta$primary_diagnosis))))))
FinalMeta$MyDiagnosis<-as.factor(FinalMeta$MyDiagnosis)
FinalMeta$censored <- FinalMeta$status


## Optional: Subset to remove gbm or just look at 1 type
FinalMeta_NoGBM<- FinalMeta[FinalMeta$project_id=="TCGA-LGG",]
FinalMeta_NoGBM$MyDiagnosis <- factor(FinalMeta_NoGBM$MyDiagnosis)
FinalMeta_Mixed<-FinalMeta[FinalMeta$primary_diagnosis=="Mixed glioma",]
```

                      
## Run univariate cox to determine which variables to control for
## RunCoxWrapper function input: 
geneSet_groups = vector of genesets names to apply wrapper to
FinalMeta = "FinalMeta"- contains time, censored, and all genesets in geneSet_groups
controlFor = whether or not this will be univariate (FALSE) or multivariate (TRUE)
mySubset = which samples used in analysis- only used for saving analysis. useful when trying out multiple subsets
exportName = analysis name to use when saving data
exportCsv = whether or not to export csv file of results
subsetToHighLow = when splitting 25/50/25, subset to just 25/25

```{r}
## Run univariate cox for all potential controlling variables
## set exportCsv to FALSE if you want to examine results before exporting
allSamples_diagnosis<- RunCoxWrapper(geneSet_groups = ToControlFor, AllMeta= FinalMeta, mySubset="AllSamples", 
                                     controlFor=FALSE, 
                                     exportName = "univariate_variables", exportCsv=FALSE)

## select just variables that had significant hazard
cox_df<-na.omit(allSamples_diagnosis$cox_df)
sig<-cox_df[cox_df$`Pr(>|z|)` < 0.1,]
sig_variables<-unique(sig$Variable)

ToBeControlled<- sig_variables
```


## Run univariate cox for genesets
## Univariate
```{r}
## Run univariate for all samples
allSamples<- RunCoxWrapper(genesets_test, FinalMeta, mySubset="AllSamples", 
                                    controlFor=FALSE, exportName = "univariate",
                                    exportCsv=FALSE, subsetToHighLow = FALSE)

## Run univariate, without gbm samples
NoGBM<- RunCoxWrapper(genesets_test, FinalMeta_NoGBM, mySubset="NoGBM", 
                                    controlFor=FALSE, exportName = "univariate",
                                    exportCsv=FALSE)


```

## Run multivariate cox, for all gene sets, controlling for variables
## Can control for 1 or more variables
## I've included some examples that were useful in my cohort, can control for whatever is appropriate for your cohort
```{r}
## Control for all sig clinical confounders- can also specify only a subset of these
allSamples_diagnosis<- RunCoxWrapper(genesets_test, FinalMeta_NoGBM,
                                     controlFor=ToControlFor,
                                     exportName="ControlForDiagnosis")

```



## Testing proportionality assumption
## Check this for at least one of the above models
## Tests whether it's appropriate to apply regression model to these variables
## If p<0.05, may need to stratify by that variable rather than controlling for it
```{r}
lapply(allSamples_diagnosis$cox_objs, function(x){cox.zph(x)})
```


## Optional: merge results together into a single df in order to summarize in a plot
Useful when you run multiple cohorts separately, but want to get an overview of all cohorts together
May need to modify for your data
```{r}
## Get all results files
allResults_files<-list.files(analysis_dir, pattern="ControlFor")
allResults_files<- allResults_files[grep("AllSamples", allResults_files)]
uni_file<-list.files(analysis_dir, pattern="univariate")

## Initiate dataframe with univariate file:
  ## Read in first file
  AllCox<- read.csv(paste0(analysis_dir, uni_file))
  
  #Subset to variables of interest, rename
  AllCox<-AllCox[,colnames(AllCox) %in% c("X", "exp.coef.", "Pr...z..")]
  colnames(AllCox)<- c("Variable", "HazardRatio", "pvalue")
  
  ## Add "ControlFor" column- "None" for univariate
  AllCox$ControlFor<- "None"

## cycle through rest of cox (multivariate), adding to univariate df
  for (i in allResults_files){
    df<- read.csv(paste0(figure_dir, i))
    df<-df[,colnames(df) %in% c("X", "exp.coef.", "Pr...z..")]
    colnames(df)<- c("Variable", "HazardRatio", "pvalue")
    
    #Get name of variable being controlled for, add to dataframe
    ControlFor<- gsub("_AllSamples_results.csv", "",i )
    ControlFor<-gsub("ControlFor", "", ControlFor)
    df$ControlFor<-ControlFor
    
    ## Remove any rows that are the variable being controlled for- not of interest
    ControlFor<-gsub("Score", "",  gsub("Primary", "Primary_", ControlFor))
    ControlFor_split<-unlist(strsplit(ControlFor, "_"))
    ControlFor_regex<- paste(ControlFor_split,collapse="|")
    df2<- df[grep(ControlFor_regex, df$Variable, invert=TRUE, ignore.case = TRUE),]
    df2<-df2[order(df2$Variable),]
    
    
    AllCox<- rbind(AllCox, df2)
  }

## Add on information about the cohort + whether gene set is categorical (high/low) or numerical (raw scores)  
AllCox$Cohort<-paste(cohort, age, sep="_")
AllCox$CategoricalOrNumerical<- grepl("Cluster", AllCox$Variable)
AllCox$CategoricalOrNumerical<- gsub("TRUE", "Categorical", 
                                     gsub("FALSE", "Numerical", AllCox$CategoricalOrNumerical))



table(AllCox$Variable)

write.csv(AllCox,paste0(analysis_dir, "allCoxResults.csv"))
```





